Imputation of Incomplete Multilevel Data with R

### 2024-01-10 Manuscript/manuscript.qmd
Update manuscript.qmd
Added link, changed affiliations and acknowledgements

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 'h'
        fig-width: 7
        fig-height: 2.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]
bibliography: bibliography.bib  
---

@@

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2) or under MNAR conditions (case study 3).

@@

### IMPACT data

@@

## Funding

@@

>>>>

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 'h'
        fig-width: 7
        fig-height: 2.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Center for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CX
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Center for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Center for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Center for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CX
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]
bibliography: bibliography.bib  
---

@@

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the background section and introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study 2, @sec-impact) or under MNAR conditions (case study 3).

@@

### IMPACT data {#sec-impact}

@@

@@

## Disclaimer
